\textsc{Beyond the Iceberg Hypothesis: \\Opening the Black Box of Transport Costs}

### 2017-07-12 redaction/dhp_july2017.tex
changement rédaction

\begin{abstract}
Following \cite{samuelson1954}, standard models of international trade have usually relied on modelling trade costs as an ad-valorem tax equivalent. However, many common empirical facts support the existence of additive costs. This paper provides a quantitative assessment of the size and the importance of the additive component in international transport costs, measured as the cif-fab price gap, using US imports flows over 1974-2013. We thus find that additive transport costs are not negligible, with mean values over the period of 1.8\% and 2.9\% of the export price in air and maritime transport respectively. Further, taking additive costs into account significantly improves the fit of the modelling of transport costs. We also use the time dimension of our data to characterize the patterns of transport costs over time. After correcting for composition effects, we find that international transport costs have decreased by 50\% in air transport and 60\% in maritime transport over the period, and of same order of magnitude for both additive and iceberg components. These results therefore confirm the importance of the additive component in accounting for international transport costs.

\textbf{pas tres sexy la fin. Dire tout de meme que ca reste non negligeable? Avec la valeur moyenne sur 2010-2013 par ex, pour additif et pour multiplicative.}
\end{abstract}

@@

\noindent Trade costs are central in international economic analysis. In particular, they are considered a major obstacle to international economic integration and international trade flows. According to \citet{Jacks08}, trade cost declines explain around 55\% of the pre-World War I trade boom and 33\% of the post-World War II trade boom, while the abrupt rise in trade costs explains the inter-war trade collapse. After 1950, average trade costs fell by 16\%, notably through the reduction of policy trade costs promoted through the GATT (WTO starting in 1995) multilateral agreements.
Based on panel data, \citet{novy13} finds  that U.S. trade costs with major trading partners declined on average by about 40 percent between 1970 and 2000.
Yet, several papers (mostly based on empirical estimates of the gravity equation) have shown that trade costs still remain a major obstacle to trade (e.g. \citealp{Head_Mayer04} and \citealp{Disdier_Head08}). Using data over 1989-2000, \citet{anderson_wincoop_jel} thus estimate that average international trade costs for industrialized countries represent a 74\% markup over production costs.

Defined as the costs associated with the exchange of goods across national borders, trade costs are usually split into transaction costs (information costs, contract enforcement costs, costs associated with the use of different currencies...), policy costs (tariff  and non-tariff costs), time costs (time to ship goods) and transport costs \emph{per se}. In this vein, \citet{anderson_wincoop_jel} obtain that around 30\% of international trade costs are attributable to transport costs. Equivalently, international transportation costs would represent a 21\% markup over production costs. \citet{Behar_Venables} obtain that the elasticity of trade with respect to freight costs is sizeable, by around -3, suggesting a non-negligible impact of transport costs on trade flows. If much of trade-policy barriers have been removed over the second half of the 20th century, these findings point out that the transport costs component of the overall trade costs remain large and deserve attention. This is accordingly the focus of the paper.\smallskip

Following \citet{samuelson1954}, standard models of international trade have usually relied on modelling trade costs as an \emph{ad-valorem} tax equivalent (ie, as a constant percentage of the producer price per unit traded, part of the ``iceberg cost'' hypothesis\footnote{Rigourously speaking, the ``iceberg cost'' includes the ad-valorem costs and the fact that the costs are paid in terms of the good that is traded. As this last element is irrelevant in our case, we use the terms ``iceberg'' and ``ad-valorem'' interchangeably, as commonly made in the literature.}). However, many common empirical facts support the existence of additive costs. As documented by \citet{Irrazabal_2015}, pricing structure in shipping, additive tariffs, distribution costs... often exhibit (at least partly) an additive structure. The structure (additive vs ad-valorem) of transport costs is far from being anecdotal, as the literature has long emphasized its role in shaping the pattern of trade flows. The Alchian and Allen conjecture (\citealp{alchian}), that points out that the relative price of two varieties of some good will depend on the level of trade costs, does rely on the existence of additive costs: The relative demand for more expensive/higher quality product goods should increase with trade cost (``shipping the good apples out''). \citet{Lashkaripour-mimeo-2016} challenges this view. He finds supporting evidence for the \emph{ad-valorem} assumption by taking into account the fact that more expensive goods are systematically heavier and hence more costly to transport. One can yet be concerned by the generality of this result. By nature, his study is restricted to goods that are enumerated by items in the statistics (they represent 60\% of US imports). Furthermore, while the positive correlation between weight and price seems reasonable for goods from the second industrial revolution like cars, it is dubious in the case of ITC goods which importance has been rising since 1994 (the end point of Lashkaripour's study). Besides, a number of empirical papers provide a strong empirical support to the role of additive costs in international costs. Based on a firm-product-level database of French exporters, \citet{martin2012} finds that firms charge higher fas unit values on exports to more remote countries, in contradiction with the ad-valorem hypothesis. \citet{hummels_skiba} estimate the elasticity of freight rates with respect to price to be well below unity. Also, their estimates imply that doubling freight costs increases average fas export prices by 80-141 percent, consistent with high quality goods being sold in markets with high freight costs. These findings deliver strong empirical support in favor of the Alchian-Allen conjecture.
\smallskip

@@

Our paper provides an empirical decomposition of the structure of transport costs over time, by explicitly distinguishing between ad-valorem and additive parts. In this respect, the paper is related to the literature that challenges the dominant role of iceberg costs in international trade.\footnote{On top of the previously cited papers about additive costs, our paper also relates to \citet{Alessandria-et-al-AER-2010}, \citet{Hornok-et-al-RES-2015} or \citet{Hornok-et-al-JIE-2015}, which point out the role of per-shipment costs (among which, administrative costs) in generating some ``lumpiness'' in international trade transactions.}
We quantitatively assess the size and the importance of the additive component in international transport costs, that we measure as the difference between the import and the export prices, based on the US Imports flows over 1974-2013. Therefore, we focus only on a subset of trade costs. Besides, in terms of transport costs \textit{per se}, our measure based on the gap between export and import prices omits the other dimension of transport costs related to the time value of goods in transit. According to \citet{anderson_wincoop_jel}, the 21\% markup over production costs coming from transport costs includes both directly measured freight costs (10.7\%) and 9\% tax equivalent of ``indirect'' costs related to the time value of goods on their way to their export market (including holding cost for the goods in transit, inventory cost due to buffering the variability of delivery dates, preparation costs associated with shipment size...). However, the evidence summarized in \citet{anderson_wincoop_jel} points to a persisting importance of direct transport costs, especially compared to other trade barriers: they remain more important than, e.g., policy barriers (8\% tax equivalent), language barrier (7\%) or information cost (6\%). \citet{Hummels_1999} mentions a bunch of papers which all point that transport costs pose a barrier similar in size, or larger than tariffs. In the same vein, \citet{limao_venables} highlight the importance of infrastructures for trade costs in general, through their impact on transport costs. Furthermore, our data is based on a single and reliable customs origin, and allows for a wide time and geographical coverage, while limiting the measurement errors coming from strategies inferring trade costs. Therefore, we strongly believe they may deliver useful and reliable insights on the structure of international trade costs.

Closely related to our paper is the work by \citet{Irrazabal_2015}, which develops a structural framework for inferring additive trade costs from firm-level trade data. Based on Norwegian exports in 2004, they find that additive costs are about 14\% of the median consumer price. Our paper complements their findings in many respects. While they study trade costs in general, our database implies that we focus on international transport costs. Similarly to them, our various results emphasize the important role of the additive component of international transport costs. Further, our empirical analysis allows us to provide a quantitative measure of the levels of both the iceberg and the additive trade costs. Last, we exploit exhaustive information about the imports flows of the US, over a large time span from 1974 to 2013. In this respect, our results deliver a broader view of the magnitude of additive costs in international trade over time. By exploiting the time coverage of our database, our paper is also related to the international trade literature that studies the patterns of trade costs over time, such as \cite{hummels2007} and \cite{Behar_Venables}. We also share in common with these papers to investigate the time trends of transport costs by transport mode (i.e., air or sea). Many argue that transport costs have substantially decreased with technological advance in transportation, infrastructure development and new communication technologies (see\citealp{Lafourcade_Thisse}). \cite{Glaeser04} find that, over the twentieth century, the cost of moving goods have declined by over 90\% in real terms. However, \cite{hummels2007} shows that the bulk of price declines in transportation comes from air transport, where average cost per ton-kilometer shipped dropped by 92\% between 1955 and 2004. Concerning ocean shipping, which represents the major part of world trade, decline in trade prices are much less obvious, a conclusion in accordance with the studies reviewed by \cite{Behar_Venables}. Our paper contributes to this debate. In particular, we show the importance of taking into account the additive component in characterizing the time trends of international transport costs. Our findings can be summarized in three main points.

First, our theoretically agnostic approach provides a fairly simple framework for measuring both ad-valorem and additive parts of transportation costs. We thus obtain that the mean values over 1974-2013 of iceberg costs are equal to 2.5\% and 3.2\% of the export price in air and ocean transport respectively, whereas the additive component amounts to 1.8\% and 2.9\%. To our best knowledge, our paper is the first to provide such an extensive quantitative measure of both ad-valorem and additive costs in total transport costs.\footnote{Given their database and their methodology, \citet{Irrazabal_2015} can only identify the ratio of the additive to the ad-valorem cost ($\frac{t_{ik}}{\tau_{ik}}$ in our terminology), that they interpret by expressing it in terms of the median export price by country-product ($\bar{\widetilde{p}}_{ik}$). By contrast, our estimation strategy enables us to uncover both values of the ad-valorem and the additive costs $\tau_{ik}$ and $t_{ik}$ separately, in terms of the country-product export price $\widetilde{p}_{ik}$.}  This represents a valuable insight for calibrating related models.

\todo{mettre une phrase pour dire qu'on confirme la litterature sur le role des additifs, mais de maniere differente. Hummels et Skiba sur des donnes de 1994, on va beaucoup plus loin qu'eux dans cette voie.}

@@

Third, we exploit the time dimension of our database to document the patterns of transport costs over time. As pointed out by \cite{hummels2007}, we confirm the importance of the composition effects by country of origin and by product in shaping the patterns of transportation costs over time. In contrast to his work though, our empirical analysis allows for a more flexible estimation of the share between the additive and the ad-valorem components of transport costs. Precisely, we allow it to change over time and across country partners and products, while \citet{hummels2007} imposes it to be constant in these three dimensions. After excluding the composition effects, we find that all types of transport costs have been roughly constant from 1974 to 1984 and then steadily decreased by 40\% over the period 1984-2013. Further, our results point out an important bias in the time trends of transport costs when additive costs are omitted. While transport costs seem to exhibit a larger decrease in air than in sea transport between 1985 and 2005 when only iceberg costs are modeled, this difference vanishes when additive costs are taken into account. These results therefore confirm the importance of the additive component in accounting for international transport costs.\smallskip

@@

>>>>

\begin{abstract}
Following \cite{samuelson1954}, standard models of international trade have usually relied on modelling trade costs as an ad-valorem tax equivalent. However, many common empirical facts suggest the existence of additive costs. This paper provides new empirical results that support this view. To do so, we exploit information contained in the US imports flows database over 1974-2013, distinguishing by transport mode (air and ocean). Two main results emerge. First, we show the additive component of transport costs is sizeable, with mean values over the period of 1.8\% and 2.9\% of the export price in air and maritime transport respectively. Further, taking additive costs into account significantly improves the fit of the modelling of transport costs. Second, we show the importance of integrating the additive component in accounting for the time trend of international transport costs. We thus show that the decrease of international transport costs observed in the data is mostly attributable to a reduction in the \textit{pure} transport costs rather than to trade pattern composition effects. These results therefore confirm the importance of the additive component in accounting for international transport costs.

\end{abstract}

@@

\noindent Trade costs are central in international economic analysis. In particular, they are considered a major obstacle to international economic integration and international trade flows. According to \cite{Jacks08}, the trade cost decline explains around 55\% of the pre-World War I trade boom and 33\% of the post-World War II trade boom, while the abrupt rise in trade costs explains the inter-war trade collapse. After 1950, average trade costs fell by 16\%, notably through the reduction of policy trade costs promoted through the GATT (WTO as off 1995) multilateral agreements.
Based on panel data, \citet{novy13} finds  that U.S. trade costs with major trading partners declined on average by about 40 percent between 1970 and 2000.
Yet, several papers (mostly based on empirical estimates of the gravity equation) have shown that trade costs still remain a major obstacle to trade (\citealp{Head_Mayer04} and \citealp{Disdier_Head08}). Using data over 1989-2000, \citet{anderson_wincoop_jel} thus estimate that average international trade costs for industrialized countries represent a 74\% markup over production costs.

Defined as the costs associated with the exchange of goods across national borders, trade costs are usually split into transaction costs (information costs, contract enforcement costs, costs associated with the use of different currencies...), policy costs (tariff  and non-tariff costs), time costs (time to ship goods) and transport costs \emph{per se}. In this vein, \citet{anderson_wincoop_jel} obtain that around 30\% of international trade costs are attributable to transport costs. Equivalently, international transportation costs would represent a 21\% markup over production costs. Further, the sizeable elasticity of trade with respect to freight costs obtained by \cite{Behar_Venables}, by around -3, suggests a non-negligible impact of transport costs on trade flows. If much of trade-policy barriers have been removed over the second half of the 20th century, these findings point out that the transport costs component of the overall trade costs remain large and deserve attention. This is accordingly the focus of the paper.\smallskip

Following \citet{samuelson1954}, standard models of international trade have usually modelled trade costs as an \emph{ad-valorem} tax equivalent (ie, as a constant percentage of the producer price per unit traded, part of the ``iceberg cost'' hypothesis)\footnote{Rigourously speaking, the ``iceberg cost'' includes both the ad-valorem dimension of trade cost and the fact that these costs are paid in terms of the good that is traded. As this last element is irrelevant in our case, we use the terms ``iceberg'' and ``ad-valorem'' interchangeably, as commonly made in the literature.} However, many common empirical facts support the existence of additive costs (i.e., cost per unit exported). As documented by \citet{Irrazabal_2015}, pricing structure in shipping, additive tariffs, distribution costs... often exhibit (at least partly) an additive structure. The structure (additive vs ad-valorem) of transport costs is far from being anecdotal, as the literature has long emphasized its role in shaping the pattern of trade flows. As pointed out by the seminal paper of \cite{alchian}, the relative price of two varieties of some good will depend on the level of trade costs as long as they display an additive component. In this context, the relative demand for more expensive/higher quality product goods increases with trade costs (``shipping the good apples out''). \citet{Lashkaripour-mimeo-2016} challenges this view. He finds supporting evidence for the \emph{ad-valorem} assumption by taking into account the fact that more expensive goods are systematically heavier and hence more costly to transport. One can yet be concerned by the generality of this result. By nature, his study is restricted to goods that are enumerated by items in the statistics (they represent 60\% of US imports). Furthermore, while the positive correlation between weight and price seems reasonable for goods from the second industrial revolution like cars, it is dubious in the case of ITC goods which importance has been rising since 1994 (the end point of Lashkaripour's study). Besides, a number of empirical papers provide a strong empirical support to the role of additive costs in international trade costs. Based on a firm-product-level database of French exporters, \citet{martin2012} finds that firms charge higher fas unit values on exports to more remote countries, in contradiction with the ad-valorem hypothesis. \citet{hummels_skiba} estimate the elasticity of freight rates with respect to price to be well below unity. Also, their estimates imply that doubling freight costs increases average fas export prices by 80-141 percent, consistent with high quality goods being sold in markets with high freight costs. These findings deliver strong empirical support in favor of the Alchian-Allen conjecture.
\smallskip

@@

Our paper provides an empirical decomposition of the structure of transport costs over time, by explicitly distinguishing between ad-valorem and additive parts. In this respect, the paper is related to the literature that challenges the dominant role of iceberg costs in international trade.\footnote{On top of the previously cited papers about additive costs, our paper also relates to \citet{Alessandria-et-al-AER-2010}, \citet{Hornok-et-al-RES-2015} or \citet{Hornok-et-al-JIE-2015}, which point out the role of per-shipment costs (among which, administrative costs) in generating some ``lumpiness'' in international trade transactions.}
We quantitatively assess the size and the importance of the additive component in international transport costs, that we measure as the difference between the import and the export prices, based on the US Imports flows over 1974-2013. Obviously, exploiting this database is not immune of limitations, in particular as it restricts our scope of study to the monetary dimension of international transport costs. Our measure being based on the gap between export and import prices, it necessarily omits the other dimension of transport costs related to the time value of goods in transit. According to \citet{anderson_wincoop_jel}, the 21\% markup over production costs coming from transport costs includes both directly measured freight costs (10.7\%) and 9\% tax equivalent of ``indirect'' costs related to the time value of goods on their way to their export market (including holding cost for the goods in transit, inventory cost due to buffering the variability of delivery dates, preparation costs associated with shipment size...). However, the evidence summarized in \citet{anderson_wincoop_jel} points to a persisting importance of direct transport costs, especially compared to other trade barriers: They remain more important than, e.g., policy barriers (8\% tax equivalent), language barrier (7\%) or information cost (6\%). \citet{Hummels_1999} mentions a bunch of papers which all point that transport costs pose a barrier similar in size, or larger than tariffs. In the same vein, \citet{limao_venables} highlight the importance of infrastructures for trade costs in general, through their impact on transport costs. Furthermore, our data is based on a single and reliable customs origin with a wide time and geographical coverage, and it allows us extracting a direct transport cost measure thereby limiting the measurement error bias. Therefore, we strongly believe exploiting this database may deliver useful and reliable insights on the structure of international trade costs.
Closely related to our paper is the work by \citet{Irrazabal_2015}, which develops a structural framework for inferring additive trade costs from firm-level trade data. Based on Norwegian exports in 2004, they find that additive trade costs are about 14\% of the median consumer price. If our study covers a narrower set of trade costs focused on transport costs, we reach a similar conclusion to them, that points out the important role of the additive component of international transport costs. Our work differentiates from them though, in two main aspects. First, our empirical analysis allows us to provide a quantitative measure of the levels of both the iceberg and the additive trade costs. Second, we exploit exhaustive information about the imports flows of the US over a large time span from 1974 to 2013. In this respect, our results deliver a broader view of the magnitude of additive costs in international trade over time.

By exploiting the time coverage of our database, our paper also relates to a second strand of the international trade literature, that studies the patterns of trade costs over time, such as \cite{hummels2007} and \cite{Behar_Venables}. Many argue that transport costs have substantially decreased with technological advance in transportation, infrastructure development and new communication technologies (see \citealp{Lafourcade_Thisse}). \cite{Glaeser04} find that, over the twentieth century, the cost of moving goods have declined by over 90\% in real terms. However, \cite{hummels2007} points out the importance of the composition effects in the characterization of time trends of international transport costs. He thus obtains a stronger decline in transportation costs between 1974 and 2004 in both air and ocean transport when the composition effects by origin country and by product have been removed. Our paper contributes to this debate. In particular, we show the importance of taking into account the additive component in characterizing the time trends of international transport costs, as it substantially modifies the decomposition of the trend reduction of transport costs between the one attributable to trade composition effects and the reduction in the ``pure'' transport costs. \textbf{le dire autrement/ mieux}

Our findings can be summarized in three main points.

First, our theoretically agnostic approach provides a fairly simple framework for measuring both ad-valorem and additive parts of transportation costs. We thus obtain that the mean values over 1974-2013 of iceberg costs are equal to 2.5\% and 3.2\% of the export price in air and ocean transport respectively, whereas the additive component amounts to 1.8\% and 2.9\%. To our best knowledge, our paper is the first to provide such an extensive quantitative measure of both ad-valorem and additive costs in total transport costs.\footnote{Given their database and their methodology, \citet{Irrazabal_2015} can only identify the ratio of the additive to the ad-valorem cost ($\frac{t_{ik}}{\tau_{ik}}$ in our terminology), that they interpret by expressing it in terms of the median export price by country-product ($\bar{\widetilde{p}}_{ik}$). By contrast, our estimation strategy enables us to uncover both values of the ad-valorem and the additive costs $\tau_{ik}$ and $t_{ik}$ separately, in terms of the country-product export price $\widetilde{p}_{ik}$.}  This represents a valuable insight for calibrating related international trade models.

@@

Third, we exploit the time dimension of our database to document the patterns of transport costs over time. Between 1974 and 2013, we thus find that transport costs have decreased by roughly 50\% in air transport and 60\% in maritime transport, and of same order of magnitude for both additive and iceberg components. Importantly, and in contrast to \cite{hummels2007}, we do not find that composition effects do make a substantial difference in the time trends of international transport costs. 
These results therefore confirm the importance of the additive component in accounting for international transport costs.\smallskip

\textbf{paragraphe precedent a reprendre}

@@

\textbf{stop here}
